---
title: "Movie Recommendation System Project"
author: "Vijayita Gumber"
date: "2024-04-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installation of required packages

```{r}
# List of packages need to be installed
packages <- c("ggplot2", "dplyr", "tidyr", "recommenderlab", "data.table", "lubridate")

for (pkg in packages) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  }
}

```
```{r}
movies <- read.csv("/Users/vgumber/Desktop/Data Science/Project/movies.csv", stringsAsFactors = FALSE)

ratings <- fread("/Users/vgumber/Desktop/Data Science/Project/ratings.csv")

```

```{r}

ratings$timestamp <- as.POSIXct(ratings$timestamp, origin="1970-01-01", tz="UTC")

ratings$year <- year(ratings$timestamp)

movies$year <- as.numeric(sub(".*\\((\\d{4})\\)$", "\\1", movies$title))

movies$genres <- strsplit(as.character(movies$genres), "\\|")

```
```{r}
print(head(ratings$timestamp))
print(head(ratings$year))
print(head(movies$year))
print(head(movies$genres))
```

## Data Exploration


Count the number of ratings given each year
```{r}
ratings$year <- as.POSIXlt(as.POSIXct(ratings$timestamp, origin="1970-01-01"))$year + 1900
yearly_rating_count <- aggregate(rating ~ year, data=ratings, FUN=length)

plot(yearly_rating_count$year, yearly_rating_count$rating, type="l", 
     xlab="Year", ylab="Number of Ratings", 
     main="Yearly Rating Count")
```

Average Ratings per year
```{r}
yearly_average_rating <- aggregate(rating ~ year, data=ratings, FUN=mean)
ggplot(yearly_average_rating, aes(x=year, y=rating)) +
  geom_line() +
  labs(x="Year", y="Average Rating", title="Average Rating Per Year")
```

No of ratings per user
```{r}
user_rating_count <- aggregate(rating ~ userId, data=ratings, FUN=length)
ggplot(user_rating_count, aes(x=rating)) +
  geom_histogram(bins=30) +
  labs(x="Number of Ratings", y="Count of Users", title="User Rating Histogram")
```

Top 10 Rated Movies
```{r}
top_rated_movies <- aggregate(rating ~ movieId, data=ratings, FUN=length)

top_rated_movies <- top_rated_movies[order(-top_rated_movies$rating), ][1:10, ]

top_rated_movies_with_titles <- merge(top_rated_movies, movies, by="movieId")

barplot(top_rated_movies_with_titles$rating, 
        names.arg = top_rated_movies_with_titles$title,
        main = "Top 10 Rated Movies",
        xlab = "Movie Title",
        ylab = "Rating Count",
        las = 2,
        ylim = c(0, max(top_rated_movies_with_titles$rating) * 1.1),  # Adjust ylim for better visualization
        cex.names = 0.4)  # Adjust the size of the names

par(las=2)
```

Distribution of Movie Rating
```{r}
avg_movie_ratings <- aggregate(rating ~ movieId, data=ratings, FUN=mean)

ggplot(avg_movie_ratings, aes(x=rating)) +
  geom_histogram(binwidth = 0.1) +
  labs(x="Average Rating", y="Count of Movies", title="Distribution of Movie Ratings")
```


Ratings Over Time for Top Movie
```{r}
top_movie_ratings <- subset(ratings, movieId == top_rated_movies$movieId[1])
ggplot(top_movie_ratings, aes(x=as.Date(as.POSIXct(timestamp, origin="1970-01-01")), y=rating)) +
  geom_line(stat="summary", fun.y=mean) +
  labs(x="Year", y="Average Rating", title="Ratings Over Time for Top Movie")

```

Correlation Between Rating Count and Average Rating
```{r}
movie_stats <- merge(avg_movie_ratings, top_rated_movies, by="movieId")
cor(movie_stats$rating.x, movie_stats$rating.y)
```

## Working on Model
```{r}
rating_data <- dcast(ratings, userId ~ movieId, value.var = "rating", fun.aggregate = mean)
rating_matrix <- as(rating_data, "realRatingMatrix")
rating_matrix
recommender_model <- Recommender(rating_matrix, method="UBCF")
model_info <- getModel(recommender_model)
summary(model_info$similarity)
```


```{r}
sampled_data<- sample(x = c(TRUE, FALSE),
                      size = nrow(rating_matrix),
                      replace = TRUE,
                      prob = c(0.8, 0.2))
training_data <- rating_matrix[sampled_data, ]
testing_data <- rating_matrix[!sampled_data, ]
```


```{r}
recommendation_system <- recommenderRegistry$get_entries(dataType ="realRatingMatrix")
recommendation_system$IBCF_realRatingMatrix$parameters
```


```{r}
recommen_model <- Recommender(data = training_data,
                          method = "IBCF",
                          parameter = list(k = 30))
recommen_model
```

```{r}
model_info <- getModel(recommen_model)

class(model_info$sim)
```
## Evaluation

Evaluation of model using RMSE
```{r}
library(recommenderlab)

set.seed(123)  # for reproducibility

users <- sample(c(TRUE, FALSE), nrow(rating_matrix), replace=TRUE, prob=c(0.8, 0.2))
training_data <- rating_matrix[users, ]
testing_data <- rating_matrix[!users, ]

model <- Recommender(training_data, method="UBCF")

predictions <- predict(model, testing_data, type="ratings")

actual_ratings <- as(testing_data@data, "matrix")

rmse <- sqrt(mean((predictions@data - actual_ratings)^2, na.rm = TRUE))
print(paste("RMSE: ", rmse))


```

Evaluating the model with personalized recommendations
```{r}
# the total no of movies recommended to each user
top_recommendations <- 10 
predicted_recommendations <- predict(object = recommen_model,
                          newdata = testing_data,
                          n = top_recommendations)
predicted_recommendations


mae <- mean(abs(predictions@data - actual_ratings), na.rm = TRUE)
print(paste("MAE: ", mae))


if (!is.vector(predictions@data)) {
  predictions_vector <- as.vector(predictions@data)  
} else {
  predictions_vector <- predictions@data
}

if (!is.vector(actual_ratings)) {
  actual_vector <- as.vector(actual_ratings)  
} else {
  actual_vector <- actual_ratings
}

valid_indices <- !is.na(predictions_vector) & !is.na(actual_vector)
predictions_filtered <- predictions_vector[valid_indices]
actual_filtered <- actual_vector[valid_indices]

confusion_matrix <- table(Predicted = predictions_filtered, Actual = actual_filtered)

print(confusion_matrix)

confusion_df <- as.data.frame.table(confusion_matrix)
colnames(confusion_df) <- c("Predicted", "Actual", "Frequency")


confusion_plot <- ggplot(confusion_df, aes(x = Actual, y = Predicted, fill = Frequency)) +
  geom_tile(color = "white") +  
  geom_text(aes(label = Frequency), vjust = 1.5, color = "black", size = 5) +  # Add text labels
  scale_fill_gradient(low = "white", high = "steelblue") +  # Color gradient
  labs(title = "Confusion Matrix", x = "Actual Label", y = "Predicted Label") +  # Labels and title
  theme_minimal()  

print(confusion_plot)
```

